A Formalization of Typed Lua

### 2015-06-13 doc/papers/paper/paper.tex
DLS: several changes

\begin{figure}[t]
\label{fig:newpoint}
\begin{verbatim}
  function new_point(x, y)
    local point = {}
    point.x = x
    point.y = y
    function point:move(dx, dy)
      self.x = self.x + dx
      self.y = self.y + dy
      return self
    end
    return point
  end
\end{verbatim}
\caption{Building an object piece-by-piece}
\end{figure}

@@

\begin{figure}[t]
\label{fig:over}
\begin{verbatim}
  function idiv(x, y)
    if y == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end

  local q, r = idiv(a, b)
  if q then
    assert(a == q*b + r)
  else
    error(r)
  end
\end{verbatim}
\caption{Overloading on multiple returns}
\end{figure}

@@

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \mathbf{void} & \textit{void type}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; T \times P & \textit{pair types}
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

@@

\begin{figure}[t]
\label{fig:subtuples}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Pair}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
	\senv \vdash P_{1} \subtype P_{2}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\begin{array}{c}
\mylabel{S-Variadic}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
{\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-VarExpand}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
	\senv \vdash F_{1}{*} \subtype P_{2}}
{\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VarDrop}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
	\senv \vdash P_{1} \subtype F_{2}{*}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]
\caption{Subtyping for second-level types}	
\end{figure}

@@

\begin{figure}[t]
\label{fig:subtable}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TableFCtoC}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..n \; \exists j \in 1..m \\
	\senv \vdash K_{j} \subtype K_{i}' \;\;\;
	\senv \vdash K_{i}' \subtype K_{j} \;\;\;
	\senv \vdash V_{j} \subtype_{c} V_{i}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{fixed|closed} \subtype\\
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}
	\end{array}} \; m \ge n
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoC}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \; \forall j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoC}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \; \forall j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoUOF}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \\
	\exists j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{unique} \subtype\\
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{unique|open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoOF}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..m \\
	\exists j \in 1..n \;
	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \in 1..n \; \not\exists i \in 1..m \;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{K_{1}{:}V_{1}, ..., K_{m}{:}V_{m}\}_{open} \subtype\\
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableFtoF}\\
\dfrac{\begin{array}{c}
	\forall i \in 1..n \; \exists j \in 1..n \\
	\senv \vdash K_{j} \subtype K_{i}' \;\;\;
	\senv \vdash K_{i}' \subtype K_{j} \;\;\;
	\senv \vdash V_{j} \subtype_{c} V_{i}'
	\end{array}}
{\senv \vdash \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{fixed} \subtype
	\{K_{1}'{:}V_{1}', ..., K_{n}'{:}V_{n}'\}_{fixed}}
\end{array}
\end{array}
\]
\caption{Subtyping for table types}
\end{figure}

\begin{figure}[t]
\label{fig:subfield}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FieldClsMtoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
	\senv \vdash V_{2} \subtype V_{1}}
{\senv \vdash V_{1} \subtype_{c} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldClsItoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldClsMtoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-FieldUnqMtoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{u} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqItoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldUnqItoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqMtoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldOpnNtoM}\\
\dfrac{\senv \vdash \Nil \subtype V}
{\senv \vdash \Nil \subtype_{o} V}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldOpnNtoI}\\
\dfrac{\senv \vdash \Nil \subtype V}
{\senv \vdash \Nil \subtype_{o} \Const \; V}
\end{array}
\end{array}
\]
\caption{Subtyping for fields}
\end{figure}

Figure~\ref{fig:subtable} shows the subtyping rules for
table types. These rules use the auxiliary relations
$\subtype_{c}$, $\subtype_{u}$, and $\subtype_{o}$
shown in Figure~\ref{fig:subfield} to handle depth
subtyping where applicable.

@@

\begin{figure}[t]
\label{fig:subrec}
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Amber}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
{\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-Assumption}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
{\senv \vdash x_{1} \subtype x_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-UnfoldRight}\\
\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
{\senv \vdash T_{1} \subtype \mu x.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UnfoldLeft}\\
\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
{\senv \vdash \mu x.T_{1} \subtype T_{2}}
\end{array}
\end{array}
\]
\caption{Subtyping for recursive types}
\end{figure}

@@

\begin{figure*}[!ht]
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s & \textit{recursive function} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}T{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}] & \textit{table assignment}\\
	& | \; id[k] \; {<}T{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}T} \; | \;
	\overline{id{:}T},{...}{:}T & \\
	\end{array}
	$$
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure*}

>>>>

\begin{figure}[t]
\begin{verbatim}
  function new_point(x, y)
    local point = {}
    point.x = x
    point.y = y
    function point:move(dx, dy)
      self.x = self.x + dx
      self.y = self.y + dy
      return self
    end
    return point
  end
\end{verbatim}
\caption{Building an object piece-by-piece}
\label{fig:newpoint}
\end{figure}

@@

\begin{figure}[t]
\begin{verbatim}
  function idiv(x, y)
    if y == 0 then
      return nil, "division by zero"
    else
      return x // y, x % y
    end 
  end

  local q, r = idiv(a, b)
  if q then
    assert(a == q*b + r)
  else
    error(r)
  end
\end{verbatim}
\caption{Overloading on multiple returns}
\label{fig:over}
\end{figure}

@@

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \mathbf{void} & \textit{void type}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{K:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; T \times P & \textit{pair types}
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Pair}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
	\senv \vdash P_{1} \subtype P_{2}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2} \times P_{2}}
\end{array}
\begin{array}{c}
\mylabel{S-Variadic}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \cup \Nil}
{\senv \vdash F_{1}{*} \subtype F_{2}{*}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-VarExpand}\\
\dfrac{\senv \vdash F_{1} \cup \Nil \subtype F_{2} \;\;\;
	\senv \vdash F_{1}{*} \subtype P_{2}}
{\senv \vdash F_{1}{*} \subtype F_{2} \times P_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-VarDrop}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \cup \Nil \;\;\;
	\senv \vdash P_{1} \subtype F_{2}{*}}
{\senv \vdash F_{1} \times P_{1} \subtype F_{2}{*}}
\end{array}
\end{array}
\]
\caption{Subtyping for second-level types}	
\label{fig:subtuples}
\end{figure}

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TableFCtoC}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \;\
	\senv \vdash K_{j}' \subtype K_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{K:V}\}_{fixed|closed} \subtype
	\{\overline{K^\prime:V^\prime}\}_{closed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{K:V}\}_{unique} \subtype
	\{\overline{K^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoC}\\
\dfrac{\begin{array}{c}
	\forall i \; \forall j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\senv \vdash \{\overline{K:V}\}_{open} \subtype
	\{\overline{K^\prime:V^\prime}\}_{closed}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableUtoUOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{u} V_{j}' \\
	\forall j  \; \nexists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{K:V}\}_{unique} \subtype
	\{\overline{K^\prime:V^\prime}\}_{unique|open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableOtoOF}\\
\dfrac{\begin{array}{c}
	\forall i \; \exists j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \land \senv \vdash V_{i} \subtype_{c} V_{j}' \\
	\forall j \; \nexists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \to \senv \vdash \Nil \subtype_{o} V_{j}'
	\end{array}}
{\begin{array}{c}
	\senv \vdash \{\overline{K:V}\}_{open} \subtype
	\{\overline{K^\prime:V^\prime}\}_{open|fixed}
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-TableFtoF}\\
\dfrac{\begin{array}{c}
	\forall j \; \exists i \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \;
	\senv \vdash K_{j}' \subtype K_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\forall i \; \exists j \;\;\;
	\senv \vdash K_{i} \subtype K_{j}' \;
	\senv \vdash K_{j}' \subtype K_{i} \;
	\senv \vdash V_{i} \subtype_{c} V_{j}'\\
	\end{array}}
{\senv \vdash \{\overline{K:V}\}_{fixed} \subtype
	\{\overline{K^\prime:V^\prime}\}_{fixed}}
\end{array}
\end{array}
\]
\caption{Subtyping for table types}
\label{fig:tablesub}
\end{figure}

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FieldClsMtoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
	\senv \vdash V_{2} \subtype V_{1}}
{\senv \vdash V_{1} \subtype_{c} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldClsItoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldClsMtoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-FieldUnqMtoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{u} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqItoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldUnqItoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqMtoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldOpnNtoM}\\
\dfrac{\senv \vdash \Nil \subtype V}
{\senv \vdash \Nil \subtype_{o} V}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldOpnNtoI}\\
\dfrac{\senv \vdash \Nil \subtype V}
{\senv \vdash \Nil \subtype_{o} \Const \; V}
\end{array}
\end{array}
\]
\caption{Subtyping for fields}
\label{fig:subfield}
\end{figure}

Figure~\ref{fig:tablesub} shows the subtyping rules for
table types. These rules use the auxiliary relations
$\subtype_{c}$, $\subtype_{u}$, and $\subtype_{o}$
shown in Figure~\ref{fig:subfield} to handle depth
subtyping where applicable.

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Amber}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
{\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-Assumption}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
{\senv \vdash x_{1} \subtype x_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-UnfoldRight}\\
\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
{\senv \vdash T_{1} \subtype \mu x.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UnfoldLeft}\\
\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
{\senv \vdash \mu x.T_{1} \subtype T_{2}}
\end{array}
\end{array}
\]
\caption{Subtyping for recursive types}
\label{fig:subrec}
\end{figure}

\begin{figure*}[t]
	$$
	\begin{array}{rlr}
	s ::= & & \textsc{statements:}\\
	& \;\; \mathbf{skip} & \textit{skip}\\
	& | \; s_{1} \; ; \; s_{2} & \textit{sequence}\\
	& | \; \overline{l} = el & \textit{multiple assignment}\\
	& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
	| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \textit{control flow}\\
	& | \; \mathbf{local} \; \overline{id{:}F} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \textit{variable declaration}\\
	& | \; \mathbf{rec} \; id{:}F = f \; \mathbf{in} \; s & \textit{recursive function} \\
	& | \; \mathbf{return} \; el & \textit{return} \\
	& | \; \lfloor a \rfloor_{0} & \textit{application with no results}\\
	& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \textit{method declaration}\\
	e ::= & & \textsc{expressions:}\\
	& \;\; \mathbf{nil} & \textit{nil}\\
	& | \; k & \textit{other literals}\\
	& | \; id & \textit{variable access}\\
	& | \; e_{1}[e_{2}] & \textit{table access}\\
	& | \; {<}F{>} \; id & \textit{type coercion}\\
	& | \; f & \textit{function declaration}\\
	& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
	| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \textit{table constructor}\\
	& | \; e_{1} + e_{2} \;
	| \; e_{1} \; {..} \; e_{2} \;
	| \; e_{1} == e_{2} \;
	| \; e_{1} < e_{2} & \textit{binary operations}\\
	& | \; e_{1} \;\&\; e_{2} \;
	| \; e_{1} \; \mathbf{and} \; e_{2} \;
	| \; e_{1} \; \mathbf{or} \; e_{2} & \textit{binary operations}\\
	& | \; \mathbf{not} \; e \;
	| \; \# \; e & \textit{unary operations} \\
	& | \; \lfloor me \rfloor_{1} & \textit{expressions with one result}\\
	l ::= & & \textsc{left-hand values:}\\
	& \;\; id_{l} & \textit{variable assignment}\\
	& | \; e_{1}[e_{2}] & \textit{table assignment}\\
	& | \; id[k] \; {<}V{>} & \textit{type coercion}\\
	k ::= & & \textsc{literal constants:}\\
	& \;\; \mathbf{false} \; | \;
	\mathbf{true} \; | \;
	{\it int} \; | \;
	{\it float} \; | \;
	{\it string} & \\
	el ::= & & \textsc{expression lists:}\\
	& \;\; \overline{e} \; | \;
	\overline{e}, me & \\
	me ::= & & \textsc{multiple results:}\\
	& \;\; a & \textit{application}\\
	& | \; {...} & \textit{vararg expression}\\
	a ::= & & \textsc{applications:}\\
	& \;\; e(el) & \textit{function application}\\
	& | \; e{:}n(el) & \textit{method application}\\
	f ::= & & \textsc{function declarations:}\\
	& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
	pl ::= & & \textsc{parameter lists:}\\
	& \;\; \overline{id{:}F} \; | \;
	\overline{id{:}F},{...}{:}F & \\
	\end{array}
	$$
	\caption{Abstract Syntax}
	\label{fig:syntax}
\end{figure*}

@@

\begin{figure*}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
\overline{\env_1, \penv \vdash e_2: fix(V), \env_v}
	\;\;\;
	\overline{\env_v, \penv \vdash e_1: K, \env_f}
	\\
	T = \{\overline{K{:}V}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{K:V} \}_{unique}\\
	\env_{1}, \penv \vdash k:K^{new}, \env_{2} \;\;\;
	\overline{K^{new} \not\lesssim K} \;\;\;
	V^{new} = fix(V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[k] {<}V{>}:V^new,
	\env_{2}[id \mapsto \{ \overline{K{:}V}, K^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdReadClose}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdReadFix}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:fix(F), \env[id \mapsto fix(F)]}
\end{array}
\end{array}
\]
\caption{Rules for Table Types}
\label{fig:tablerules}
\end{figure*}
